% Copyright (c) Antti Järvinen 2022. See licensing section for details,
% shortly CC-BY-SA. For html or pdf output you'll need latex package
% hyperref to have external links working.
\documentclass{article}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[colorlinks = true,
            linkcolor = blue,
            urlcolor  = blue,
            citecolor = blue,
            anchorcolor = blue]{hyperref}

% define hyperlink command with colours:
\newcommand{\chref}[3][blue]{\href{#2}{\color{#1}{#3}}}%

\begin{document}

\title{Between a rock and a hard place}
\author{Antti J\"arvinen \\ antti.jarvinen@katiska.org}
\date{July 2022}

\maketitle

\begin{abstract}
Everything you read in here is most likely all wrong, see
\chref{https://en.wikipedia.org/wiki/Karl_Popper}{Karl Popper} for details.

Forms of life are units able to make copies of themselves. Co-operation between life-forms
makes the co-operative unit to become life-form it its own. In Earth the result is multiple
layers of co-operating life-forms, where humans are in one layer. Understanding the benefits
of co-operation and constraints set by adjacent layers should help humans to decide how to
co-operate efficiently without sacrificing friendliness, sustainability, happiness or love. 
\end{abstract}

\section{Licensing of this article}
This article is copyright \copyright Antti Järvinen 2022. 
This article may be distributed according to limitations of
\chref{https://creativecommons.org/}{creative commons}
license \chref{https://creativecommons.org/licenses/by-sa/4.0/}{Attribution-ShareAlike} that
roughly gives every recipient of this work right to copy and mix contents of this article as
they wish. Copies and works deriving from this article must give credit to original
author. Copies and derivative works can be used commercially but it must be made clear
to recipients of those copies and derivate works that same licensing still apply e.g. copies
may be given away free of charge and the derivative work may be used in yet another
derivate work under same license. PDF version of
\chref{http://katiska.org/~costello/between.pdf}{this document is available here}.

\section{Life does not exist}
Ordinary matter, things that we can touch, things that are heavy to lift, things that
we feel as breeze in our skin while we ski against wind in freezing temperature are all
made of elementary particles. This is best understanding of current state of affairs
that humankind has at the time of writing. As K. Popper in previous chapter pointed out,
this is subject to change. Everyones contribution to better understanding of the world
as it is is required and nobody should be shy of voicing his/hers opinion - spectacular
findings may have already been found, thing just might be that nobody as enough courage
to speak them out. Making funny assumptions about world or life is no crime, if they
turn out to be false then it is ok. That is how science actually works - there are
assumptions that are proven false. Those assumptions that are not proven false after
some time are considered "maybe true". This is the best that science can achieve.

Ordinary matter is made of elementary particles
\chref{https://en.wikipedia.org/wiki/Up_quark}{Up},
\chref{https://en.wikipedia.org/wiki/Down_quark}{Down} and
\chref{https://en.wikipedia.org/wiki/Electron}{Electron}. Up and Down quarks join together
to form sub-atomic particles \chref{https://en.wikipedia.org/wiki/Proton}{Proton} and
\chref{https://en.wikipedia.org/wiki/Neutron}{Neutron}. Protons and neutrons together with
elementary particle electron join together in partially understood rules to produce
\chref{https://en.wikipedia.org/wiki/Atom}{atoms} whose
stable combinations are listed in \chref{https://en.wikipedia.org/wiki/Periodic_table}{periodic table} of atoms.

The "forces" that glue quarks into sub-atomic particles are also particles, named
\chref{https://en.wikipedia.org/wiki/Boson}{Bosons}.

The "force" that glues sub-atomic particles into atoms is electromagnetic attraction
between atomic nuclei and electrons.

Atoms can get arranged in structures called
\chref{https://en.wikipedia.org/wiki/Molecule}{Molecules} and they in turn are
held together by chemical bonds, largely dictated by behavior and number of electrons
of atoms in molecule.

Observation here is that, matter has "layers", each layer having different
constituents and interactions. 
\begin{table}[h!]
\begin{center}
\begin{tabular}{l|l|l}
\textbf{Constituents} & \textbf{Forces} & \textbf{Result of layer} \\
\hline
Quarks & Bosons & Sub-atomic particles \\
Sub-atomic particles + Electron & Electromagnetic attraction & Atoms \\
Atoms & Chemical bonds & Molecules \\
\end{tabular}
\caption{Layers of particles that ordinary matter is made of}
\end{center}
\end{table}
While it can be argued that "everything is made of elementary particles only" it is
also very useful to identify these layers - when smaller particles are grouped together
to produce a "2nd level" particle the "new" particle starts having different kind
of interactions between other particles in the same layer. For atoms be behavior
between atoms is largely dictated by electrons in outermost electron shell. Also new
properties may emerge: states like gaseous, liquid and solid may be identified in
atoms but make little sense if applied to fundamental particles. 
\subsection{Chemistry}
\chref{https://en.wikipedia.org/wiki/Chemistry}{Chemistry} as a science studies how atoms
join together to produce molecules.
It studies which combinations of atoms are possible in different environments
where temperature, pressure, number of different atoms present etc. varies.
Molecules have new properties like possible un-even distribution of electric
charge and 3-dimensional shape. Shape and electric charge affect how different
kind of molecules "fit together" and interact together - interaction here
means transferring of electron from one molecule to another or molecules may break
and their atoms get rearranged into yet new molecules - these are called chemical
reactions. 

As molecules as new layer on top of atom-layer adds new properties and new interactions
that don't occur with "plain" atoms, each layer still must obey rules of all underlying
layers. With molecules it means that molecules still must be made of atoms found
from periodic table and atoms must be mostly made of up+down quarks and electrons. 

Important concept in chemical reactions is
\chref{https://en.wikipedia.org/wiki/Catalysis}{catalysis} where catalyst is a substance
(a molecule) that, when present in the soup where chemical reaction occurs the catalyst
changes the reaction. Catalyst has such a shape or electric distribution that it makes
other molecules in the soup to fit together in different ways compared to situation
where catalyst is not present. Catalyst can change speed of reaction (how many molecules
are created or split apart in one second) or yield of the reaction (how many \% of starting
material molecules gets converted into resulting molecules). Humans design catalytic
molecules for different purposes but many naturally occuring molecules have catalytic
properties.

\section{Autocatalytic molecules}
\chref{https://en.wikipedia.org/wiki/Autocatalysis}{Autocatalytic} reaction is important
"special case" of catalysis. Molecule is autocatalytic if it can catalyze reaction
of itself. Simply, if there are building blocks of autocatalytic molecule around and
one autocatalytic molecule, presence of autocatalytic molecule will cause building
blocks (atoms, other molecules) to arrange into similar molecule as the original autocatalytic
one. It can be understood that autocatalytic molecule can produce a copy of itself
given that there are required building blocks available. As autocatalytic process
advances, also number of resulting autocatalytic molecules increases and each molecule acts as a
catalyst, speed of autocatalytic reaction could increase until shortage of starting
materials starts to make reaction not possible any more.

Nature of chemical reactions is that usually not 100\% of staring materials is
converted in reaction and also there may be several slightly different resulting
molecules. For example many organic molecule building blocks fit together so that
resulting molecule is twisted either to left or right side. Mixture containing
both left+right organic molecules is called
\chref{https://en.wikipedia.org/wiki/Racemic_mixture}{racemic mixture}. Another possible
source of variation is situation where starting material building block (atom or other
molecule) is substituted with another substance that is still chemically close enough
to "usual" building block. Example might be that
\chref{https://en.wikipedia.org/wiki/Lithium}{lithium} is one building block of
molecule. It happens that no lithium atom is available during reaction but 
\chref{https://en.wikipedia.org/wiki/Sodium}{sodium} atom is. Lithium and sodium
have similar electron configuration in outer shell and may substitute each
others in some chemical reactions. Simple examples are lithium chloride and sodium chloride.
Following observations must be made
\begin{itemize}
\item Lithium chloride and sodium chloride are different chemicals and have
      different behavior towards other materials.
\item Both lithium atom and sodium atom will happily create a bond
      with single chloride atom, because electron configurations are similar enough.
\end{itemize}
\subsection{Life does exist}
Autocatalytic reaction is no different from any chemical reaction ; there
may be variation in resulting molecules, in other words not all copies are exactly
like original. Shape of molecule may be different, some atoms and other building blocks
may have been replaced with "compatible" parts. Because molecule is then different, it will
also react differently. Differences may make it stronger or weaker catalyst. 
If new building block was introduced and that particular building block is available
in environment in large quantities, it may enable continuing the autocatalytic reaction
also after all original building blocks were already used. In autocatalytic reaction having
diverse set of starting materials around the following happens:
\begin{itemize}
\item Copies are made,
\item Not all copies are similar,
\item Infinite number of copies is not possible.
\end{itemize}
this leads to situation where "best" configurations of autocatalytic molecule are found
in mixture in greatest concentrations. "Best" may be measured for example by following
traits:
\begin{itemize}
\item Is strong catalyst e.g. speed and yield of reaction is high. High does not
      need to be high in absolute terms. It just needs to be higher than any
      "competiting" configuration of autocatalytic molecule.
\item Molecule is stable, e.g. it does not break easily.
\item It is made only of building blocks that are available in large quantities
      in environment. 
\end{itemize}
The molecules that better than others implement the "best molecule" metrics then
appear in resulting chemical soup more often than others. This natural process of
selection is called \chref{https://en.wikipedia.org/wiki/Evolution}{evolution}
and it starts functioning always when copies of something are made, not all copies
are similar and there may be only finite number of those copies in existence.

Currently best known autocatalytic molecule is \chref{https://en.wikipedia.org/wiki/DNA}{DNA}
that is found inside cells of all organisms that people tend to call "living". DNA
is result of evolution. It has ancestors. Most likely those ancestors are no more,
all material (atoms, small molecules) that those ancestors were made of have been used
to build next generations of the ancestor - they may be present inside modern configuration
of DNA but are difficult to trace. DNA is super complex molecule. Its early ancestors
must have started small. Studies have been made about what kind of "original" autocatalytic
molecule could have been, for example in
\chref{https://www.eurekalert.org/news-releases/939921}{Rutgers university}
and
\chref{https://www.frontiersin.org/articles/10.3389/fmicb.2021.793664/full}{Düsseldorf and Vienna}. 

\section{Forces of interaction}
In terms of layers and interactions between layers the list of layers now looks something like
in table 2:
\begin{table}[h!]
\begin{center}
\begin{tabular}{l|l|l}
\textbf{Constituents} & \textbf{Forces} & \textbf{Result of layer} \\
\hline
Quarks & Bosons & Sub-atomic particles \\
Sub-atomic particles + Electron & Electromagnetic attraction & Atoms \\
Atoms & Chemical bonds & Molecules \\
Molecules & Chemical bonds & More complex molecules \\
Complex molecules & Chemical bonds & Autocatalytic molecules\\
\end{tabular}
\caption{Layers of particles that ordinary matter is made of}
\end{center}
\end{table}

Autocatalytic molecules from table above don't necessarily need to be complex but DNA is.
In basic interactions the "molecule" layer adds some new properties to interactions
between different molecules, like shape and electric charge and distribution of electric
charge between different parts of the molecule. "Autocatalytic molecule" above is
presented as layer of its own but basically it is just molecule in chemical reaction.
It has anyway one new interesting property and that the number of copies it can
make of itself in every possible environment. Here environment means mixture of
different chemicals, for example sea water or less salty river water. 

\section{Viruses, cells and single-celled organisms}
Today modern DNA produces not only copies of itself but also contains
molecular mechanism for producing number of supporting molecules, called
\chref{https://en.wikipedia.org/wiki/Protein}{proteins} that make
autocatalysis of DNA more safe, error-free and enables it to happen
in more diverse environments. Current understanding is that first
\chref{https://en.wikipedia.org/wiki/Viral_evolution}{viruses} appeared
about same with \chref{https://en.wikipedia.org/wiki/Cell_(biology)}{cellular
organism} and it is difficult to say which was first. Today
viruses require a cell to replicate themselves. Cell in turn is
support mechanism that DNA has building instructions for. Situation today
is that all identified complex lifeforms are made of cells. There are
life forms consisting of only one cell - DNA is packed safely inside
protecting \chref{https://en.wikipedia.org/wiki/Cell_membrane}{cell membrane}
and some \chref{https://en.wikipedia.org/wiki/Unicellular_organism}{single
celled organism} have DNA for instructions to grow
\chref{https://en.wikipedia.org/wiki/Flagellum}{hair-like} appendices
for swimming around. Thing to note here is: \emph{cell is support mechanism
for DNA to replicate itself} in environment where other organism might
utilitize (say: eat) DNA building blocks to make copies of itself and
where finding of all required building blocks for DNA replication requires
a little bit of swimming around.

Comparing to previously presented list of "best molecule" metrics
the \chref{https://en.wikipedia.org/wiki/Protein}{protein} makes sense
very much: \chref{https://en.wikipedia.org/wiki/Enzyme}{enzymes} are
class of proteins that break down other proteins and also
\chref{https://en.wikipedia.org/wiki/Deoxyribonuclease}{DNA} itself.
Any DNA configuration that has developed mechanism for disassembling
DNA building blocks from nearby organisms is in great advantage in
making copies of itself compared to DNA configurations that don't
contain instructions for such behavior. Notable thing here is that
\emph{behavior is determined by laws of physics and chemistry}. DNA
strain that is complex molecule can not have functions that humans
call "thinking", "seeking of goal" or "conscience", instead its
survival is determined by how well it implements the "best molecule"
practices. Those practices may be different in different environments
or vary from time to time but it is merely a chemical process with no
conscience that drives evolution of DNA into direction where resulting
DNA implements environmental requirements in best way. And "best"
is measured by number of DNA copies, only. It doesn't necessarily
mean "more complex", "more friendly to nearby DNA constructs" or
"more bound to develop proteins that code Frank Zappa\footnote{Not even in 
the night of the iron sausage} music in understandable way". 

Viruses today are "free floating" DNA snippets that require living cell
to replicate. They anyway contain instructions for their own replication,
often with some \chref{https://en.wikipedia.org/wiki/Influenza}{side-effects}
for the cell doing the replication. Viruses have evolution of their own,
they compete in how to best infect cells to make them produce copies
of the virus. From perspective of this paper viruses are entity that
is capable of copying functional DNA sequences from one living cell to
another e.g. copying instructions for codeing some protein ( == complex
molecule having specific chemical properties) from one life-form to
another. In single-celled organism the addition or replacement to
DNA of the cell is immediately inherited to descendants of the
single-celled organism. 

When continuing the above-presented table of "layers of ordinary matter"
here comes a jump. Previously layers for most part were constuctions of
particles from underlying layer. Here, next layer is constucted from
information it carries using order of particles from underlying layer.

\begin{table}[h!]
\begin{center}
\begin{tabular}{l|l|l}
\textbf{Constituents} & \textbf{Forces} & \textbf{Result of layer} \\
\hline
Quarks & Bosons & Sub-atomic particles \\
Sub-atomic particles + Electron & Electromagnetic attraction & Atoms \\
Atoms & Chemical bonds & Molecules \\
Molecules & Chemical bonds & More complex molecules \\
Complex molecules & Chemical bonds & Autocatalytic molecules (DNA)\\
Complex molecules like amino acids of DNA & Order of amino acids & Proteins\\
Proteins and simple molecules & Protein structure & Cells + single cell organisms\\
\end{tabular}
\caption{Layers of particles that ordinary matter is made of}
\end{center}
\end{table}

in table 3 above the "forces" column lists the source of information
how next layer is constructed but there is more: cells and single-cell
organisms have ways of interacting with each other also after their
constuctions on complete. Single-cell organism interact with numerous
ways. One way is predator-catch interaction where one organism consumes
another as food. More complex interactions are detailed in
\chref{https://www.sciencedirect.com/science/article/pii/S2452310021000081}
{article} "The landscape of cell–cell communication through single-cell 
transcriptomics" where information transfer between single-cell organism is
believed to be mostly using biochemical signaling e.g. exchange of various
molecules.

If previously signaling or information exchange between particles of
ordinary matter have been mostly defined by mass(==frequency) of
particle, here comes a fundamental change: information between
single-cell organisms is exchanged in
form of order of elements (like amino acids) or order of atoms (causing
specific 3-dimensional shape of molecule that affects ways how subsequent
chemical reactions proceed).
\section{Multi-celled organisms}
Multi-celled organism is similar to single-celled but with (at least)
following additional features:
\begin{itemize}
\item DNA has instructions for producing different kind of cells for
      different purposes. Each cell has copy of same "build instructions"
      in form of DNA but there is \chref{https://en.wikipedia.org/wiki/Hox_gene}
      {mechanism} to selectively use only parts of DNA for each cell
      depending on its purpose. "Purposes" for humans include cells
      for skin, blood and brain etc. that are all different, human
      body has around 40 different issue types coded in DNA.
\item There may be specific system for communication between cells inside
      multi-celled organism called
      \chref{https://en.wikipedia.org/wiki/Nervous_system}{nervous system}
      that sends information between other cells. System consists of
      cells whose sole purpose is to send information between other cells.
      Information here means things like "fingertip is too hot", which
      in successful case would translate into muscle cell nervous command
      telling the finger muscles to get fingertip away from heat source.
\item Alone cell that is differentiated for purpose is usually not
      able to live without its friends that are coded in the same DNA.
      This is stark differience compared to single-celled organism.
      Each single-celled organism can live and make copies of itself
      by its own means. Individual multi-celled organisms can't live or
      make copies of themselves if not all different cell types coded
      in its DNA are present and in correct places in its body. Many
      multi-celled organisms have specific cell types only for sharing the DNA,
      for example \chref{https://en.wikipedia.org/wiki/Sperm}{sperm cell}
      and \chref{https://en.wikipedia.org/wiki/Egg_cell}{egg cell}
      that together contain enough DNA to produce fertile decendants.
      Methods in so-called sexual re-production where DNA from both
      parents is required has actually many variations, for example there
      are animals and plants
      capable of \chref{https://en.wikipedia.org/wiki/Heterogamy}{heterogamy}
      or animals like ants that don't divide DNA equally
      between \chref{https://en.wikipedia.org/wiki/Ant\#Reproduction}
      {father and mother}.      
\end{itemize}
Important thing to notice here is that \emph{multicellular organisms
communicate with other multicellular organisms using means and language
that single-celled organisms don't have.} Multi-cellular organisms
have specific cell-types for communication, like cell-types for
\chref{https://en.wikipedia.org/wiki/Hearing}{hearing} and
\chref{https://en.wikipedia.org/wiki/Visual_perception}{seeing}
and some cell-types do not participate at all what it comes
to communication with other cells around. All this is not so simple,
there is still plenty of single-celled organisms around like
\chref{https://en.wikipedia.org/wiki/Bacteria}{bacteria} that will try
to communicate with each and every cell around, including cells of
multi-celled animal. Compound of cells, that is called "multi-cell organism"
starts to behave like single entity, having its own communication methods
and maybe also reproductive goals that maybe are partly focused
to re-production of single cell but also focusing in re-productive
success of the whole, whole here being collection of DNA strains
for each and every cell type in the multi-cell organismn. 

There is evolution of single-cell organism DNA but with multi-cell
organism it is the DNA of the multi-cell organism (like human) that
gets replicated as whole, so the superset of DNA of all differentiated
cells in multi-cell organisms becomes unit evolution drives
to "best molecule" objectives. "Best molecule" objectives become much
more complicated here. Some set of DNA of multi-cell organism may have
superior set of building instructions for escaping predators, other set may have
superior set of building instructions for finding food
and all this in comparison to other organism in varying environments.

\begin{table}[h!]
\begin{center}
\begin{tabular}{l|l|l}
\textbf{Constituents} & \textbf{Forces} & \textbf{Result of layer} \\
\hline
Quarks & Bosons & Sub-atomic particles \\
Sub-atomic particles + Electron & Electromagnetic attraction & Atoms \\
Atoms & Chemical bonds & Molecules \\
Molecules & Chemical bonds & More complex molecules \\
Complex molecules & Chemical bonds & Autocatalytic molecules (DNA)\\
Complex molecules like amino acids of DNA & Order of amino acids & Proteins\\
Proteins and simple molecules & Protein structure & Cells + single cell organisms\\
Single cell organisms & Nervous system & Multi-cell organisms\\
\end{tabular}
\caption{Layers of particles that ordinary matter is made of}
\end{center}
\end{table}

In table above only "nervous system" is listed as communicating
force between single cells and members of multi-celled organism.
It is not only one but it is the most important. Some "single cells"
differentiate to become carriers of information between other cells. Result
is that multi-celled organism can have objectives shared between different
cell types like muscle, blood and nervous. Here happens same thing
when sub-atomic particles form an atom - it starts to communicate
with other atoms in different means than sub-atomic particles do.
Here happens same thing when atoms group together to form molecules -
molecules start to interact with other molecules in methods that
are unknown or alien to single atoms. Here happens same thing
where constituents of complex autocatalytic molecule form
single-cell organism - single cell organism starts to communicate and
have goals (or interpretation of "best molecule" practices) that
has wider objectives than structure of autocatalytic molecule
that does not contain instructions for cell membrane or other
"utilitiary functions" a cell may have.

Groups start to behave like single entity. This concerns communication
between entities in same layer but also common reproductive goals.

\section{Groups of multi-celled organisms}
Some single-cell organisms form groups that exhibit group behavior. Example
is \chref{https://en.wikipedia.org/wiki/Dental_plaque}{dental plaque}
that seems behave differently depending on number of its own
species in neighbourhood. Measurement is maybe done using
amount of biofilm it generates.

Some multi-celled organisms are not teamplayers. Most still require
partner for mating so at least temporary team-play is a requirement.
Team-play doesn't necessarily need to be friendly in terms of
well-being of multi-cellular organism, like in
\chref{https://www.nationalgeographic.com/animals/article/spiders-mothers-cannibals-arachnids}{this example}
that in turn may be beneficial for species if number of DNA copies
in the environment is counted. Organism, is it single-cell or multi-cell
is always extension to DNA that is used to produce copies of
autocatalytic DNA molecule. In DNA terms "success" is counted
only as number of similar DNA strains in the vicinity. Wasp is not
a wasp. Wasp is vehicle that DNA-strain called "Wasp DNA" uses to
make more copies of wasp DNA strains. This vehicle has many virtues
like building a nest, flying around, protecting offspring and
finding the best queen to mate with. Carrot is not a carrot. Carrot is
vehicle that DNA-strain called "Carrot DNA" uses to make more copies
of carrot DNA strains. This vehicle has many virtues like producing
seeds, looking eatable to humans so they'll carry seeds around for
planting purposes etc. 

\subsection{Language}
Some forms of communication from earlier "layers of particles that ordinary
matter is made of" can be considered methods of group communication. Example is
electromagnetic attraction that at least slightly "visible" to all
electronically charged molecules nearby. Topic of this chapter anyway
is multi-cellular organisms that have dedicated cells or cell structures
for communication between other organisms, let it be same species or
other. These structures include organs like hands for making gestures,
eyes for seeing those gestures, vocal cords for producing voices,
ears for hearing those voices and neural cells capable of learning
writing and reading of alphabet so that written communication becomes
possible. 

Complex multicellular organisms have complex nervous system capable of
remembering, observing and imitating. Combining imitation and remembering
enables \emph{learning} from other group members. Evolution of language
as feature of multicellular organism must have started with expressions
like calls of warning or mating. In humans and maybe some other
vertebrates ability to imitate gestures and sounds has led to development
of \chref{https://en.wikipedia.org/wiki/Language}{languages} that may be
spoken, \chref{https://en.wikipedia.org/wiki/American_Sign_Language}{signed}
with gestures, written and
\chref{https://en.wikipedia.org/wiki/Internet_protocol_suite}{transmitted}
electronically around the globe, just to give a few examples. Before
emergence of \emph{general purpose language capable of describing
any phenomena or event} each individual could remember only things
it had witnessed in person. Language enables sharing of memories, enables
individuals from group take into considerations descriptions (let them
be false or true) of events and things originally witnessed by other
group members, maybe long time ago. Result is \emph{group knowledge}
that can include more facts and skills than any individual group
member with limited brain capacity could alone handle, know and use
in decision-making. Ideas or \chref{https://en.wikipedia.org/wiki/Meme}{memes}
mediated by language can directly contribute DNA replication of individual
who remembers and acts out an idea, for example porridge making.
Individuals who don't know how to make porridge may be at disadvantage
compared to porridge-making individuals in terms of arranging DNA
building blocks required for DNA replication. Other ideas like
\chref{https://en.wikipedia.org/wiki/Birth_control}{birth control}
may have also impact on DNA replication. Yet other ideas like
usage of \chref{https://en.wikipedia.org/wiki/Fidget_spinner}{fidget spinner}
maybe less so. Humans are not agnostic to these ideas. Those ideas
that are useful like porridge-making, funny like good jokes or ideas
that trigger powerful emotions like awe, fear or love are more likely
to be re-told to other group members. These ideas have following properties
\begin{itemize}
\item Humans, maybe other vertebrates too, make copies of ideas by
      sharing them using language. Copies are stored in memory of
      each individual.
\item Copies are not identical because humans sometimes describe
      things incorrectly, understand incorrectly and also over time
      tend to remember facts incorrectly. Humans also sometimes
      make new discoveries like "novel method for peeling 200
      potatoes under one minute using bare hands only" that are
      then re-told and humans also make blatant lies for various
      purposes like "it is possible to peel 200 potatoes in one minute".
\item There is only limited number of memory locations (== individual
      heads with brain and memory inside) available.
\end{itemize}
Result is similar selection of \emph{best ideas} in
\chref{https://en.wikipedia.org/wiki/Evolution}{evolutionary process} as
happens with autocatalytic molecules. Unit that is selected is
an idea or story, something that can be told or shown or imitated.
Where landscape of autocatalytic molecules is a chemical soup, here
landscape of replicating ideas is language, physiology and psychology
of group members. "Best idea" metrics could include things like
\begin{itemize}
\item Is so useful to individual that it gets re-told, like methods
      of building hunting weapons.
\item Makes lasting memory, like idea that touching broken electric
      wire may cause electric shock and that may trigger fear of
      injury or death.
\item Triggers or mediates some psychological need, like willingness
      to be loved or desire to control behavior of other group members. 
\end{itemize}
List above is longer. Language itself is an idea subject to "best idea"
metrics and languages do indeed evolve. While chemical evolution is
mostly random, evolution of ideas has additional strong "intentional" component
in it. Language if not enables, at least greatly facilitates
\chref{https://en.wikipedia.org/wiki/Herd_behavior} {herd behavior}.

\subsection{Herd behavior}
In biological evolution the unit that is selected is DNA inside cells
of individual, this applies also to herd members. Uniform or compatible
behavior between group members may still contribute to DNA strains
chances of getting copied. Examples of beneficial group behaviors
could be fending off an attacker together or maintaining hygiene
together for getting rid of a disease. Human is herd-animal too.
Many herds or groups of humans are such that members of the group are
not equal in terms of capabilities and responsibilities between
group members - different \chref{https://en.wikipedia.org/wiki/Pack_hunter}
{roles between group members} predate emergence of general purpose
language but language must make definition of different roles inside group
more precise, if not easier. For successful co-operation group members
must share at least partially compatible set of ideas in their memory.
Modern herds are formed around geographical locations, biological relatedness,
profession or subject of interest ; anything where human can "belong to".
Typical human belongs to multiple herds at the same time, for example
via citizenship, employment and membership to a sports club. Some
of these organizations are such that organization members share common
ideas about roles, privileges and responsibilities - most notably
nation-states often have concept of written "law" that may define
roles like "foreign minister", "educator" or "social worker" and
define some of their responsibilities inside organization. Notable thing
here is that language facilitates this behavior. Another notable thing is
that these groups develop means of communications and interactions
between other similar groups. Humans understand and sometimes accept this
notion. Examples are statements like "Inter Milan football club wins
Serie A" - statement does not list individual athletes in the club,
instead club is an entity that plays, does, wins, loses, gives statements.
Communication like scoring a goal or objecting to referees judgement
still happens between humans but they're understood to re-present the
group, organization or herd. The group may have obligations towards its
members, like salary, right to education or health care, group members may have
obligations towards group, like military service. Group behavior
gives emergence of concepts that make no sense between single human
individuals only, for example "foreign debt" or "economic zone".
\emph{Group of individuals may start behaving like single entity,
these entities evolve their own language and concepts that are relevant
to group-entities.}

\begin{table}[h!]
\begin{center}
\begin{tabular}{l|l|l}
\textbf{Constituents} & \textbf{Forces} & \textbf{Result of layer} \\
\hline
Quarks & Bosons & Sub-atomic particles \\
Sub-atomic particles + Electron & Electromagnetic attraction & Atoms \\
Atoms & Chemical bonds & Molecules \\
Molecules & Chemical bonds & More complex molecules \\
Complex molecules & Chemical bonds & Autocatalytic molecules (DNA)\\
Complex molecules like amino acids of DNA & Order of amino acids & Proteins\\
Proteins and simple molecules & Protein structure & Cells + single cell organisms\\
Single cell organisms & Nervous system  & Multi-cell organisms\\
Multi-cell organisms & Language, ideas & Tight group-behavior \\
Individuals in group & Inter-group concepts & Actions between groups \\
\end{tabular}
\caption{Layers of particles that ordinary matter is made of}
\end{center}
\end{table}

Table of layers has now conceptional jumps, and it should be divided
into 2 directions. One direction is presented above, language knits
individuals into groups that have some internal organization and order,
tight or loose. Language can also be separating element, for example
speakers if different languages may feel alien to other residents
of same geographical area that speak some other language. In "group
concepts" single individual is not topic of discussion any more than
single muscle cell in a healthy human individual. Human individual is usually
referenced as a whole, including all its cells with DNA inside. Same
happens here, humans have developed a concept where "Inter Milan wins"
or "Sweden develops a new vaccine" with no need to list individual
athletes or scientists.

Second "direction" in table above would be to explore how ideas get grouped
together to form "group of ideas" that maybe get communicated e.g. copied
together. This is discussed in next chapter. 

\section{Culture}
Some ideas come in groups. Origin and evolution of some groups of ideas is
\chref{https://en.wikipedia.org/wiki/Intellectual_history}{subject
to scientific research}. Human mind mostly works so that badly contradicting
ideas can't be considered true at the same time. Statements like "entrance
to building is found behind a corner", "all buildings are round", "every
building has an entrace" make together little sense so their chances
of getting re-told together might be small. They're 3 individual ideas anyway,
but together make up only nonsense. Some groupings would make sense,
like "method for planting potatoes", "calendar for determining best time
to plant potatoes", "cookbook of most delicate servings of 200 potatoes"
that could be part of idea-group called "agriculture".

\chref{https://en.wikipedia.org/wiki/Culture}{Culture} can be almost
any kind of grouping if grouped ideas make sense together. Or idea
doesn't need to "make sense" - example is rainbow colors that some individuals
considering themselves as part of LGBT culture tend to display. It can
be argued that rainbow features every color of visible light spectrum
and is that way suitable symbol for that particular culture. Selection
of symbol is anyway arbitrary, instead depiction of selected animal
or flower would do same job in identifying members of the culture.
Raibow is currently used, it may change in the future. Currently
"set of ideas" that are contained under umbrella "LGBT culture" comes
with a rainbow symbol because some early activists started to use that.
Once in use, usage tends to stick until there is good reason to replace it.

\section{Cultural evolution}
Cultures evolve as ideas inside culture evolve. Cultures may also
acquire new ideas to be included in the culture. For example
fish-eating population having rich culture regarding catching and
preparing fish but knowing nothing about potatoes comes across an
individual carrying kettle of boiled potatoes. Should they purposefully
or accidentally mix boiled potatoes with fish, result might be
new cultural item "fish soup with potatoes" that soon would be in
minds of almost every member of fish-eating population, of course
not counting the conservative puritans that say that "fish is a fish
and should be eaten just as it is".

New additional idea to culture needs to have some appeal to members
of the culture. It may be taste, like in fish-soup. Or useful skill
like counting days from calendar. Or promise that fills a psychological
need like "after death martyrs go to heaven". Surely all non-trivial
cultures include some ideas that are true and some ideas that are not true.
Lies may be included because they can, especially if they are hard
to prove false and if they trigger some strong emotion in humans,
making the whole culture appear more strongly in individuals mind.
Many cultures that define nation-states or religions include stories
that include depictions of death, love, fraud or trust etc., items that
trigger strong emotions in most individuals. Not every idea inside
a culture needs to be true, or not blatantly exaggerated to be useful
for the culture.

Biological evolution and multi-cellular organisms often deploy some sort
of \chref{https://en.wikipedia.org/wiki/Immune_system}{immune system}
that is supposed to keep invading or alien DNA from being copied
in cells of the organism. Cultural evolution doesn't have anything
very sophisticated for the same purpose. Racism towards individuals
identified to entertain another culture may be one adaptation into that
direction. It doesn't target ideas, instead it targets individuals
with the idea. Different languages may also act as "natural barriers"
between some cultures. Nowadays common ability for humans to understand
more than one language makes this barrier leak quite a bit. 

\section{Conscience and limitations of biology and culture}
Evolution of cultures is a must-have thing, unstoppable process.
Every bigger culture that author is aware of has some ideas inside that make
human life miserable if acted out. And some ideas that make human life good
if acted out. Culture as set of stories to act out as human is very defining
aspect of human life for many individuals. Many humans exercise very little
selection over individual ideas in a culture they feel being part of and act out
stories and beliefs in real life without much consideration. 
\subsection{Between a rock..}
Human as a regular multicellular organism is constrained by its building
instructions that it carries coded in DNA inside every cell. DNA replication
is automatic process, it happens when required building blocks are in
place and laws of chemistry feel no mercy. Building blocks must be
gathered, environmental conditions in body like temperature and
acidicity must be maintained or DNA-copying vehicle called "human body"
will gradually stop working. Chemical process like replication of DNA
doesn't have concepts of "good" or "evil", reaction will be triggered
when required starting materials are in place. This is how autocatalytic
molecules work. Some behaviors like willingness to seek food, avoid
predators and produce offspring are coded into DNA - humans don't need
to learn to seek food, it happens naturally, usually parents help a bit
with techniques \emph{how to seek food} but human baby will try to seek food
naturally from day one after it is born. Many
\chref{https://en.wikipedia.org/wiki/Big_Five_personality_traits}{personality
traits} are partly inherited, partly dictated by environment and
childhood experiences. That in turn is largely guided by something 
called "culture". 
\subsection{..and a hard place}
Just like DNA-copying chemical reaction has no conscience, culture as a
set of ideas is not self-ware, it does not think without help from human,
it can be considered a virus a the mind that affects behavior of its
host, usually including behavior that spreads the culture to more
minds and tries to restrict competeting cultures from reaching the
already-infected mind.
\subsection{Of conscience and science}
In "layers of particles that ordinary matter is made of" the layer
"multi-cellular organisms" is one and only that has processes implementing
self-awareness, self-reflection, notion of good (using metrics like "to whom",
"when", "at what cost" etc.) and emotions like feeling of hope.
Chemical process has no feelings. Set of ideas has no feelings.
Humans act out requirements set by biology, ultimately goals set by
DNA replication. Humans act out requirements set by culture they feel
they belong to. Some cultures may employ psychological tricks in
infected mind, like triggering guilt or fear if elements of culture
are not acted out, for example threats of miserable afterlife.

Our own biology we can't change very much. Attempts to tamper with human
DNA can come with very unexpected side-effects and may be really difficult
to reverse if gone wrong. Humans are still not held accountable to
any large chemical. For example, if DNA is prevented from being
copied using \chref{https://en.wikipedia.org/wiki/Birth_control}{birth
control} it will not feel bad or try to retaliate. Humans don't need to
be slaves to biology but instead understand its limitations, what is possible
and what is not. This includes as well physiological limitations as
limitations of human mind that \chref{https://en.wikipedia.org/wiki/Psychology}
{psychology} studies.

Our own culture we can change as we will. Biological viruses need
a host cell to replicate, sometimes replication ends with errors
that sometimes are beneficial for the virus. Ideas and cultural items
need brain and mind to replicate, sometimes replication ends
with errors that sometimes are beneficial for the virus. The brain
and mind doing the replication have control over what is copied,
when is copied, why is copied. Humans do things, do good things,
sometimes do stupid things, any of these are candidates for imitation
and thus candidates to become part of cultural heritage.

Most humans come equipped with working conscience and some understanding
about what is good and what is not. "Good" is extremely difficult concept,
there it needs to be considered at least
\begin{itemize}
\item Good to whom - self or someone else?
\item When, immediately or maybe later, maybe during long period of time
\item What are consequences of not implementing the good thing
\end{itemize}

Humans, as a race or individual are not held accountable to a set of
ideas. If some ideas in the set get modified, replaced or forgotten
the culture is not going to feel bad or try to retaliate. Some cultures
come with concept of "law" or "football club rule of order" that may
contain rewards or punishments to individuals depending on how they act
out the cultural ideas. Humans should not consider themselves to be
slaves of their culture but instead consider each and every idea
pending implementation in light of conscience, measure its goodness
in relation to self and also other living organisms, living now and
coming after us.

Together with conscience should in parallel be applied
\chref{https://en.wikipedia.org/wiki/Science}{science} that includes
measuring, validating, testing hypothesis and building knowledge.
Repeatable measurements and findings are more likely to be true.
If culture itself defines something to be true or false, scrutiny
may be in order, information may be outdated and some cults
even encourage members of the cult for lying. Lying may be telling
false facts about leader-figure of the cult and upon hearing the
falsehoods, other members of the cult recognise the individual as
one who is willing to trade his reputation to membership in the cult.
Reputation e.g. trust in others that individual speaks the truth
is in most circles considered valuable. If individual is willing to sacrifice
this valuable feature by telling lies in public, he clearly values the
cult membership more than his own reputation and so may be accepted to
cult. Typical lies are claims that everybody knows as false, for example
so-called "miracles".

With knowledge of the world and its workings it may be possible to gain
a little bit of wisdom about how to act out, what to do, what to not do.
This is even more difficult than considering good and bad in light
of the conscience. Humans might need a cultural evaluation culture
for developing stories, methods, acts and rules purposefully.
This needs to be a continuous group exercise.

\end{document}
